---
title: "State-Level Police Union Contributions"
author: "Aaron Kessler"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(plotly)
library(htmltools)
library(widgetframe)
library(scales)
library(zoo)
library(lubridate)
library(gt)
library(kableExtra)
library(writexl)
options(scipen = 999)

#load the data tables from step 01
contribs_tocands <- readRDS("processed_data/contribs_tocands.rds")
contribs_topartycmtes <- readRDS("processed_data/contribs_topartycmtes.rds")
contribs_combined_candspartycmtes <- readRDS("processed_data/contribs_combined_candspartycmtes.rds")
contribs_toballotmeasures <- readRDS("processed_data/contribs_toballotmeasures.rds")
currentofficeholders_byparty <- readRDS("processed_data/currentofficeholders_byparty.rds")
statepops <- readRDS("processed_data/statepops.rds")
forbignumber <- readRDS("processed_data/forbignumber.rds")


contribs_tocands %>% 
  group_by(general_party) %>% 
  summarise(sum(dollar_amount)) 

contribs_topartycmtes %>% 
  group_by(party) %>% 
  summarise(sum(dollar_amount)) 

contribs_tocands %>% 
  group_by(incumbency_status) %>% 
  summarise(tot_dollars = sum(dollar_amount)) %>% 
  mutate(tot_dollars / sum(tot_dollars) * 100)



#if want to limit the years choose here
timeframe_vector <- c("2014", "2016", "2017", "2018", "2019", "2020")


#create per capita measures and rankings
staterankings_tocands_all <- contribs_tocands %>% 
  group_by(election_jurisdiction) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  inner_join(statepops, by = c("election_jurisdiction" = "state")) %>% 
  mutate(
    dollars_per1000ppl = (total_dollars / population) * 1000,
    rank_dollars_per1000ppl = min_rank(desc(dollars_per1000ppl)),
    rank_population = min_rank(desc(population))
  ) %>% 
  arrange(rank_dollars_per1000ppl)

staterankings_tocands_all %>% write_xlsx("output/staterankings_tocands_all.xlsx")


staterankings_topartycmtes_all <- contribs_topartycmtes %>% 
  # filter(election_year %in% timeframe_vector) %>% #filtering years yere
  group_by(election_jurisdiction) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  inner_join(statepops, by = c("election_jurisdiction" = "state")) %>% 
  mutate(
    dollars_per1000ppl = (total_dollars / population) * 1000,
    rank_dollars_per1000ppl = min_rank(desc(dollars_per1000ppl)),
    rank_population = min_rank(desc(population))
  ) %>% 
  arrange(rank_dollars_per1000ppl)

staterankings_topartycmtes_all %>% write_xlsx("output/staterankings_topartycmtes_all.xlsx")


staterankings_tocombinedcandsparties <- contribs_combined_candspartycmtes %>% 
  # filter(election_year %in% timeframe_vector) %>% #filtering years yere
  group_by(election_jurisdiction) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  inner_join(statepops, by = c("election_jurisdiction" = "state")) %>% 
  mutate(
    dollars_per1000ppl = (total_dollars / population) * 1000,
    rank_dollars_per1000ppl = min_rank(desc(dollars_per1000ppl)),
    rank_population = min_rank(desc(population))
  ) %>% 
  arrange(rank_dollars_per1000ppl)

staterankings_tocombinedcandsparties %>% write_xlsx("output/staterankings_tocombinedcandsparties.xlsx")

contribs_tocands %>% 
  count(election_year)


```


```{r narrative_vals, include=FALSE}
###this chunk used for working out R code for inserting pieces in-line into narrative below

# big number grand total for entire universe
vector_bignumber <- forbignumber %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull()


#total to Democratic cands/cmtes
vector_total_dem_candscmtes <- contribs_combined_candspartycmtes %>% 
  filter(party == "Democratic") %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull()

#total to Republican cands/cmtes
vector_total_gop_candscmtes <- contribs_combined_candspartycmtes %>% 
  filter(party == "Republican") %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull()

#total to CURRENT Democratic cands/cmtes
vector_total_dem_CURRENTcandscmtes <- currentofficeholders_byparty %>% 
  filter(general_party == "Democratic") %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull()

#total to CURRENT Republican cands/cmtes
vector_total_gop_CURRENTcandscmtes <- currentofficeholders_byparty %>% 
  filter(general_party == "Republican") %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull()


incumb_summary <- contribs_tocands %>% 
  group_by(incumbency_status) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  mutate(pct = round_half_up(total_dollars / sum(total_dollars) * 100))

vector_incumbentpct <- incumb_summary %>% filter(incumbency_status == "Incumbent") %>% pull(pct)
vector_challengerpct <- incumb_summary %>% filter(incumbency_status == "Challenger") %>% pull(pct)
vector_openseatpct <- incumb_summary %>% filter(incumbency_status == "Open") %>% pull(pct)


```

## Lede

It's been a long time coming, but recent weeks have finally brought about a sea change in public opinion toward police in the wake of George Floyd's death in Minneapolis. Police abuses of power and the targeting, and killing, of black Americans, has sent thousands of people into the streets to protest across the country on a scale not seen in decades. And surveys now show broad support nationwide for reforms to hold police accountable.

But whether this moment translates into the political will strong enough to overcome police departments' repeated ability to thrwart efforts to change the system remains to be seen. The real test will be in the states themselves, where the police and their powerful unions are major players, having spent tens of millions of dollars to curry favor with state and local politicians. 

In fact, police-affiliated unions and associations have contributed **`r vector_bignumber`** to state candidates, political parties and ballot initiatives across the United States since the 2000 election, a new Bloomberg Industry Group analysis found.

The majority of that money has gone to Democrats, who in some places have responded by blocking meaningful change to end excessive use of force, even after horrific incidents. Minnesota's inaction after the 2016 death of Philando Castile could be exhibit A. Baltimore, where Freddie Gray died in police custody in 2015 as a result of a catastrophic neck injury, similarly failed to subsequently take meaningful action.

Democratic candidates and party committees recieved **`r vector_total_dem_candscmtes`** compared with Republicans' **`r vector_total_gop_candscmtes`** over the last two decades, according to the analysis of campaign data compiled by the nonprofit National Institute on Money in Politics.

And the Democratic state politicians holding office right now at this very moment? They've collectively received about **`r vector_total_dem_CURRENTcandscmtes`** from law enforcement unions over the course of their careers -- nearly triple the **`r vector_total_gop_CURRENTcandscmtes`** of their current Republican counterparts. 

To be sure, the analysis shows police groups favored the candidates with power - those who are already *incumbents* - more so than a particular party. An overwhelming **`r vector_incumbentpct` percent** of police group money to candidates went to incumbents over the last two decades. Another  **`r vector_openseatpct` percent** of the money went to contenders for open seats, where there was no current officeholder to run against.  
Only  **`r vector_challengerpct` percent** of contributions by police groups went to candidates challenging an incumbent.

That said, it doesn't take the pressure off Democrats, who are in a bind that they TKTKTKTKTK.... It's likely that TKTKTKTK....
  
  
## Top candidates, top donors

Who are the candidates who have gotten the biggest bails of cash from police unions over the years? They include both the current Democratic governors of New York, **Andrew Cuomo**, and California, **Gavin Newsom**.

```{r, echo=FALSE}

#clean up Andrew Cuomo records to standardize his listings since they are recorded three ways
contribs_tocands <- contribs_tocands %>% 
  mutate(
    candidate = if_else(candidate == "CUOMO, ANDREW M & DUFFY, ROBERT J", "CUOMO, ANDREW M", candidate),
    candidate = if_else(candidate == "CUOMO, ANDREW M & HOCHUL, KATHLEEN COURTNEY (KATHY)", "CUOMO, ANDREW M", candidate)
  )

topcands_dem <- contribs_tocands %>% 
  filter(general_party == "Democratic") %>% 
  group_by(candidate, election_jurisdiction) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  head(10)

vector_cuomo <- topcands_dem %>% 
  filter(candidate == "CUOMO, ANDREW M") %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull(total_dollars)

vector_denapoli <- topcands_dem %>% 
  filter(candidate == "DINAPOLI, THOMAS P") %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull(total_dollars)

vector_newsom <- topcands_dem %>% 
  filter(candidate == "NEWSOM, GAVIN") %>% 
  mutate(total_dollars = dollar(total_dollars)) %>% 
  pull(total_dollars)

#display table for top dems
topcands_dem %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")


```

**Cuomo** tops the list of any currently elected state official for contributions from police groups - he's received a total of **`r vector_cuomo`** (though the arcane nature of New York campaign finance rules means some of that money was given to a joint account he shared with running mates). The longtime state comptroller of New York, **Thomas DiNapoli**, is not far behind with **`r vector_denapoli`**. California's **Newsom** has taken **`r vector_newsom`**.

(The champion overall, former California governor Gray Davis, hasn't been in office since the early 2000s; he was sent packing after an anti-establishment upheaval resulted in a 2003 recall election that scrambled California politics.)  
  
And here are the **top Republican candidates**, all but one of them New York politicans...


```{r, echo=FALSE}

topcands_gop <- contribs_tocands %>% 
  filter(general_party == "Republican") %>% 
  group_by(candidate, election_jurisdiction) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  head(10)

topcands_gop %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")


```


The **top police unions and associations** since 2000 in total giving to both parties is perhaps not surprisingly concentrated among groups in California and New York...perhaps not surprising. The states here at the top are huge.

```{r, echo=FALSE, warning=FALSE}
topdonors <- contribs_combined_candspartycmtes %>% 
  group_by(contributor) %>% 
  summarise(total_dollars = sum(dollar_amount)) %>% 
  arrange(desc(total_dollars)) %>% 
  head(10)

topdonors %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")



```

## Where the money dominates

But if we look at **state-by-state giving accounting for the populations**, we find something interesting.  

Firstly, it turns out California and New York still top the list even after adjusting for their populations.  
  
But perhaps more strikingly are the smaller states that have a *substantially larger* amount of police union contributions. Rhode Island, Arkansas, and Delaware rank **at the bottom by population**, yet they're among the **highest ranked for per-capita police contributions**.  
New Jersey and Washington State, both with well-known problems with their police departments' use of force, also have notably outsized campaign contributions as well given their lower populations.

```{r, echo=FALSE}
staterankings_tocombinedcandsparties %>% 
  head(10) %>% 
  mutate(
    total_dollars = scales::dollar(total_dollars),
    dollars_per1000ppl = scales::dollar(dollars_per1000ppl)
) %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(6, bold = T, color = "purple") %>% 
  column_spec(5, bold = T, color = "red") %>% 
  column_spec(1, bold = T, color = "black") 


```


```{r, echo=FALSE}




```


```{r, echo=FALSE}




```

